Trauma treatment delay

### 2023-12-07 manuscriptlongversion.Rmd
text

Karolinska University hospital, which is the equivalent of a trauma
level 1 hospital, admits 1500 trauma patients each year. To be added to
the trauma registry, a patient must be over 15 years of age, had ISS\>9
and/or had trauma code activation. They get added to both the Karolinska
Trauma registry as well as the national trauma registry (SweTrau). The
registry includes data on vital signs, injuries, interventions and
patient information. Another database, the trauma care quality database
includes information relevant for MoM conferences. At Karolinska
University Hospital, patients who die during hospital stay gets their
case sent to a MoM everytime. However, for the rest of the patients, two
specialized nurses determine whether they are. Several audit filters are
set up to flag for patients with possible OFIs. Flagged patients are
then reviewed by these two nurses who decide if there is a possible OFI,
in which case their case is sent to a MoM conference for further review.

@@

Selection bias as this is a retrospective cohort study where the outcome
has occurred.

Misclassification bias as the conference may reach a consensus that is
wrong.

@@

The logistical regression is run through Rstudio using the glm command.
To make the tables, the package gtsummary was used to visualize the
adjusted and unadjusted regression.

@@

In the following table, the results from the second regression analysis
is displayed.

The odds ratio is calculated in the same way as the previous regression
model, as well as a 95% confidence interval. The p value is also shown.

The analysis shows that some variables seem to have a correlation with
patients receiving delayed treatment. ISS and respiratory rate both have
a trend that patient with higher values seem to have a lower incidence
of delayed treatment, both unadjusted and adjusted. GCS had the opposite
correlation, with lower GCS often having lower incidence of delayed
treatment, both unadjusted and adjusted. Number of injuries had the
correlation of higher number of injuries having lower incidence of
delayed treatment unadjusted, but that correlation was not seen when
adjusted for.

@@

These results show that there seems to be a correlation between some
variables and the presence of delayed treatment. A higher GCS had a
correlation with delayed treatment while a lower ISS had the same
correlation. This suggests that a less injured and more awake patient
has a higher risk of receiving delayed treatment. Compared to previous a
previous study analysing the same data but with ofi as the outcome, this
result is surprising as that study came to the opposite result
[@Pat_factors]. This can possibly be explained by that the study looked
at all ofi, whereas this study concentrates on delayed treatment.
Delayed treatment might have this correlation because an injured and
unconscious patient might receive more attention at the hospital,
leading to fewer instances delays compared to a patient that seems less
injured.

Respiratory rate also had the correlation that lower respiratory rate
correlated with lower rates of delayed treatment among the patients. As
the interval is \>10 to \<29 in the triage guideline, both too low and
too high respiratory rate is not desirable[@EHCP2015]. This might have a
similar explanation to ISS and GCS, that bradypnea in a patient might
attract more attention in the hospital and lower the incidence of
delayed treatment. Another possible explanation might be that patients
with tachypnea might receive higher rates of delayed treatment because
they are more severely injured, therefore a higher risk of missing a
possible injury. Another similar study, which focused on ofi generally,
showed the same correlation, although the significance dissapeared in
the adjusted analysis.

@@

Limitations

In this study, one of the limitations was the amount of people having
the recorded outcome being quite small, leading us into having to limit
the amount of variables we wanted to examine. Furthermore, the original
plan was to divide RR, SBP and GCS into revised trauma score (RTS)
categories, from 1-4. However, many of the RTS variables had few to no
patients with the observed outcome. This led to separation in the
logistical regression in these variables, referring to when an outcome
variable separates a predictor variable completely, which occurs when
the predictor is only associated with one outcome value [@OARC]. In the
case of these variables, it was only RTS 3 and 4 who had patients with
the observed outcome, which left RTS 1 and 2 with confidence intervals
of 0.00 to infinity. This way would have been the best way to examine
these variables, but as that wasn't possible with this group of
patients, we resorted to making these variables continuous.

@@

Interpretation

Give a cautious overall interpretation of results considering
objectives, limitations, multiplicity of analyses, results from similar
studies, and other relevant evidence

@@

>>>>

Karolinska University hospital, which is the equivalent of a trauma
level 1 hospital, admits 1500 trauma patients each year. To be added to
the trauma registry, a patient must be over 15 years of age, had ISS\>9
and/or had trauma code activation. They get added to both the Karolinska
Trauma registry as well as the national trauma registry (SweTrau). The
registry includes data on vital signs, injuries, interventions and
patient information. Another database, the trauma care quality database
includes information relevant for MoM conferences. At Karolinska
University Hospital, patients who die during hospital stay gets their
case sent to a MoM everytime. However, for the rest of the patients, two
specialized nurses determine whether they are. Several audit filters are
set up to flag for patients with possible OFIs. Audit filters may
include low SBP, high RR, EMS scene time greater than 20 mins etc.
Flagged patients are then reviewed by these two nurses who decide if
there is a possible OFI, in which case their case is sent to a MoM
conference for further review.

@@

Selection bias as this is a retrospective, pots hoc cohort study where
the outcome has occurred.

Misclassification bias as the conference may reach a consensus that is
wrong, alternatively that the two specialized nurses going through the
patients flagged for audit filters might missclassify patients or that
the audit filter misses a patient with ofi.

@@

The logistical regression is run through Rstudio using the glm command.
It was ran several times to make an ajusted table and 7 unadjusted, one
for each variable To make the tables, the package gtsummary was used to
visualize the adjusted and unadjusted regression.

This process was the only time AI was used in the writing process. It
was used to make snippets of code for a specific purpose, revise and
optimize code and find solutions when written code was not functioning
properly.

@@

In the following table, the results from the logistical regression
analysis is displayed.

The odds ratio is calculated both unadjusted and adjusted for other
variables, as well as a 95% confidence interval. The p value is also
shown.

The analysis shows that some variables seem to have a correlation with
patients receiving delayed treatment. ISS and respiratory rate both have
a trend that patient with higher values seem to have a lower incidence
of delayed treatment, both unadjusted (ISS, OR 0.96, 95% CI
0.95-0.98)(RR, OR 0.95, 95% CI 0.91-0.98) and adjusted (ISS, OR 0.97,
95% CI 0.95-0.99)(RR, OR 0.96, 95% CI 0.93-0.99). GCS had the opposite
correlation, with lower GCS often having lower incidence of delayed
treatment, both unadjusted (OR 1.13, 95% CI 1.07-1.20) and adjusted (OR
1.08, 95% CI 1.00-1.15). Number of injuries had the correlation of
higher number of injuries having lower incidence of delayed treatment
unadjusted (OR 0.94, 95% CI 0.91-0.98), but that correlation was not
seen when adjusted for. The three remaining variables, age, gender and
SBP, showed no correlation with the observed outcome.

@@

These results show that there seems to be a correlation between some
variables and the presence of delayed treatment. A higher GCS had a
correlation with delayed treatment while a lower ISS had the same
correlation. This suggests that a less injured and more awake patient
has a higher risk of receiving delayed treatment. Compared to previous a
previous study analysing the same data but with ofi as the outcome, this
result is surprising as that study came to the opposite result
[@Pat_factors]. This can possibly be explained by that the study looked
at all ofi, whereas this study concentrates on delayed treatment.
Possible explanations to the correlation include the idea that
unconsious and more severely injured patients receive more immediate
attention at the hospital, leading to fewer delays.

Respiratory rate also had the correlation that lower respiratory rate
correlated with lower rates of delayed treatment among the patients. As
the interval is \>10 to \<29 in the triage guideline, both too low and
too high respiratory rate is not desirable[@EHCP2015]. This might have a
similar explanation to ISS and GCS, that bradypnea in a patient might
attract more attention in the hospital and lower the incidence of
delayed treatment. Another possible explanation might be that patients
with tachypnea might receive higher rates of delayed treatment because
they are more severely injured, therefore a higher risk of missing a
possible injury. Another similar study, which focused on ofi generally,
showed the same correlation, although the significance dissapeared in
the adjusted analysis [@Pat_factors].

@@

The fact that age and gender did not show an increased odds ratio should
be seen as a positive sign. It suggests that among trauma patients,
there is no systematical delay based on gender or age.

In this study, one of the limitations was the amount of people having
the recorded outcome being quite small, leading us into having to limit
the amount of variables we wanted to examine. Furthermore, the original
plan was to divide RR, SBP and GCS into revised trauma score (RTS)
categories, from 1-4. However, many of the RTS variables had few to no
patients with the observed outcome. This led to separation in the
logistical regression in these variables, referring to when an outcome
variable separates a predictor variable completely, which occurs when
the predictor is only associated with one outcome value [@OARC]. In the
case of these variables, it was only RTS 3 and 4 who had patients with
the observed outcome, which left RTS 1 and 2 with confidence intervals
of 0.00 to infinity. This way would have been the best way to examine
these variables, but as that wasn't possible with this group of
patients, we resorted to making these variables continuous. We
considered expanding the outcome group to include more ofi, but opted
for the solution used in the study instead as we were solely interested
in this ofi for this study.

@@

Interpretation and clinical applications

Further discussing the results and discussing it's clinical
applications, the results can be interpreted as that it is mostly
uninjured and awake patients that receive delayed treatment. The
clinical applications for this are not clear, but might suggest that
personnel working with trauma patients should be aware of that people
that seem uninjured might be more likely to receive delayed treatment
and trying to lower that risk.

@@

##
